Interisland movements of woodland caribou in Newfoundland

### 2020-02-21 paper/manuscript.Rmd
Update manuscript.Rmd, manuscript.docx, and 2 more files...

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
   ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
   ^2^Department of Biology, Memorial University of Newfoundland, Canada
   #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
  #  github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
   bookdown::word_document2:
    reference_docx: template.docx
bibliography: ../references/knitcite.bib
#csl: csl.csl
---

@@

## Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016]
to avoid predators [@Bergerud_1985] and access islands during calving
[@Bergerud_1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running, while drowning is also possible [@Miller_1986].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Jeffery_2007;
@Ricca_2012]. <!-- add Miller 1995/2002 or reference table 1 if we decide to
keep this -->

The Fogo Island archipelago, off the coast of Newfoundland, Canada,
is home to approximately 300 woodland caribou. During routine fieldwork on May
30, 2017, we observed an unmarked adult male caribou swim between Wester and
Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
(n = 29 adult females: 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket,
Ontario, Canada). We identified swimming events when two consecutive GPS (2-hour
relocation rates) locations from an individual occurred on different islands.
This region typically experiences pack ice during the winter, so caribou may be
able to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) and we
identified three adult female caribou that collectively had 86 swimming
events (Figure 2).

@@

## Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer_2015] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer_2015]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller_1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud_1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

@@

## Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
(Miller 2002) <!-- find doi for this article -->. During calving, female caribou
are known to swim to small islands in freshwater lakes give birth, or with their
calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline
habitat in summer is considered an effective anti-predator calving strategy by
allowing for swimming to islands or peninsulas as a predation avoidance flight
response [@Bergerud_1990]. The use of water, and by association occasionally
swimming, is therefore part of the fine-scale interactions between caribou and
their predators and the use of islands appears to be an effective anti-predator
strategy.

@@

## Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. 
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
ultimate explanation to a series of proximate observations, fine-scale
density-dependent habitat selection is theoretically possible
[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
habitat was strongest at low population densities, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit -->

@@

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by the National
Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
acknowledge the territory in which data were collected and analyzed as the
ancestral homelands of the Beothuk, and the island of Newfoundland as the
ancestral homelands of the Mi’kmaq and Beothuk.
<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->

## Literature Cited

>>>>

---
title: Interisland movements of woodland caribou in Newfoundland
author: Quinn M.R. Webber^1\*^, Jack G. Hendrix^1^, Alec L. Robitaille^2^,  Eric Vander Wal^1,2^
date: 
   ^1^Cognitive and Behavioural Ecology Interdisciplinary Program, Memorial University of Newfoundland
   ^2^Department of Biology, Memorial University of Newfoundland, Canada
   #\*Corresponding author: Quinn M.R. Webber (webber.quinn@gmail.com)
output: 
  #  github_document
  # bookdown::pdf_document2:
  #    number_sections: true
  #    toc: false
   bookdown::word_document2:
    #reference_docx: template.docx
bibliography: ../references/knitcite.bib
#csl: csl.csl
---

@@

# Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond_2016],
to avoid predators [@Bergerud_1985], and access islands during calving
[@Bergerud_1990]. Even for adept swimmers like caribou, the energetic
expenditure of swimming for quadrupedal mammals is significantly higher than
walking or running [@Fish_1994], while drowning is also possible [@Miller_1986].
Despite a large number of coastal and island caribou herds, only a handful of
examples have documented caribou swimming in the ocean [e.g. @Jeffery_2007;
@Ricca_2012]. <!-- add Miller 1995/2002 or reference table 1 if we decide to
keep this -->

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between Western and Eastern Indian
Islands (Figure 1), a swim of at least 470 m which took ~9 minutes (~52 m per
minute). This observation prompted us to investigate whether swimming behaviour
was common among GPS radio-collared caribou in the population (n = 29 adult
females; 1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario, Canada). We
identified swimming events as two consecutive GPS (2-hour relocation rates)
locations from an individual occurred on different islands. Northeastern
Newfoundland typically experiences pack ice during winter and caribou are known
locally to travel between islands by walking over the ice. We restricted our GPS
data to the ice-free period of the year (April 1 - November 25) and we
identified three adult female caribou that collectively had 86 swimming events
(Figure 2). In addition to our own visual and remotely sensed observations,
residents of Fogo Island have also reported observing caribou swimming between
islands on numerous occaisions.

@@

# Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
[@Schaefer_2015] and could explain movement between islands. Conspecific
competition for resources driven by reduced foraging opportunities can decrease
reproductive success of adult females [@Schaefer_2015]. Forage scarcity has been
proposed as a potential reason for caribou moving between arctic islands on the
sea-ice in winter [@Miller_1977]. For very small islands, over-grazing by newly
arrived caribou may rapidly deplete forage [@Bergerud_1990]. Without sea ice,
movement from the mainland to an island, or between islands, requires caribou to
assess the trade-off associated with swimming.

@@

# Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
(Miller 2002) <!-- find doi for this article -->. During calving, female caribou
are known to swim to small islands in freshwater lakes give birth, or with their
calves after birth, to avoid predation [@Bergerud_1987]. The use of shoreline
habitat in summer is considered an effective anti-predator calving strategy by
allowing for swimming to islands or peninsulas as a predation avoidance flight
response [@Bergerud_1990]. The use of water, and by association occasionally
swimming, is therefore part of the fine-scale interactions between caribou and
their predators and the use of islands appears to be an effective anti-predator
strategy.

@@

# Individual variation 
In total, 3 of 29 collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 <!--double check--> days during ice-free seasons. However, despite
relatively frequent swimming events for these three individuals, we were unable
to identify any swimming events for the other 26 collared females. It remains
unclear why ocean swimming is relatively common among ~10% of our marked
population, while the other 90% never swam. Of the three collared individuals,
two were collared on Western Indian Island and the third was collared on Fogo
Island. It is possible that individuals born on, or that live on, Eastern and
Western Indian Island swim are more likely to swim between islands because
inter-island distances are shorter than between Western Indian and Fogo Islands
(Figure 2). Furthermore, it is unclear whether swimming is more common among
males or females. Our initial observation  was of an adult male, but our GPS
data only includes females. 
<!-- QW: I'm not sure we need to say this.. 
It is theoretically possible that these other
females moved rapidly back and forth between islands within the 2 hour fix rate
of our GPS collars, which we would be unable to detect. We consider this highly
unlikely given the range of residency times for caribou on any given island (min
- max: XX - XX). <!--double check-->
<!-- What is the typical residency time? Is this a true statement or am I just making this up? QW: this is something we will look into, I think. -->

@@

# Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris_1987]. While this is an
ultimate explanation to a series of proximate observations, fine-scale
density-dependent habitat selection is theoretically possible
[@Webber_2017]. For horses (*Equus ferus caballus*), selection for high quality
habitat was strongest at low population densities, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory [@van_Beest_2013]. Predator avoidance could also
contribute to caribou swimming behaviour, but given the presumed low density of coyotes on Fogo Island and that predation primarily only affects young calves, we suspect predator is a possible, but less likely explanation.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible? 
QW: changed it a bit -->

@@

# Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by the National
Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
acknowledge the territory in which data were collected and analyzed as the
ancestral homelands of the Beothuk, and the island of Newfoundland as the
ancestral homelands of the Mi’kmaq and Beothuk.
<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->

# Literature Cited
